Data Exploration: IUCN and AquaMaps

### 2015-11-21 R/data_explore_pt1.Rmd
trying out new maps by quadrant to see if there are common characteristics...

---
title: 'Data Exploration: IUCN and AquaMaps'
output: html_document
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# Libraries and Paths

library(ggplot2)
library(maps)
library(readr)
library(raster)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
```

```{r read_am_and_iucn_data, echo = FALSE, warning = FALSE, message = FALSE, results = 'hide'}

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_both.csv'), 
                           col_types = 'cdddd', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) ### removes extinct subpopulations

am_spp_cells   <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells.csv'), progress = TRUE)

### species info list - scinames, IUCN species ID, AM species ID, category etc
spp_list_base <- read.csv('shiny/data/spp_am_v_iucn.csv', stringsAsFactors = FALSE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- 'shiny/data/loiczid_raster.grd'
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

``` {r ggplot_theme_setups, echo = FALSE}

### generic theme for all plots
ggtheme_plot <- theme(axis.ticks = element_blank(),
                    text = element_text(family = 'Helvetica', color = 'gray30', size = 8),
                    plot.title = element_text(size = rel(1.5), hjust = 0, face = 'bold'),
                    legend.position = 'right')

# ggtheme_map <- theme(axis.ticks = element_blank(), axis.text = element_blank(),
#                     text = element_text(family = 'Helvetica', color = 'gray30', size = 8),
#                     plot.title = element_text(size = rel(1.5), hjust = 0, face = 'bold'),
#                     legend.position = 'right')

### theme for species range maps - based on generic plot theme
ggtheme_map <- ggtheme_plot +
  theme(axis.text = element_blank())

@@

``` {r species_map_functions, echo = FALSE, warning = FALSE, message = FALSE}

### Species Map Function ###
# This function takes a single species scientific name as input, then grabs all 
# occurrence cells and associated Aquamaps probability and/or IUCN proportional area
# per cell
get_spp_map <- function(species){
  spp_id <- spp_list_base %>%
    filter(sciname == species & is.na(parent_sid)) %>%
    dplyr::select(am_sid, iucn_sid, sciname) %>%
    unique()
  iucn_spp_map <- iucn_spp_cells %>%
    filter(sciname == species) %>%
    group_by(loiczid) %>%        
    summarize(iucn_area = max(prop_area))
  ### The group_by() and summarize() are to collapse cells with overlapped polygons, fixing the duplicate 'by' issue
  am_spp_map   <- am_spp_cells %>%
    filter(am_sid == spp_id$am_sid) %>%
    rename(am_prob = prob)
  spp_map <- full_join(iucn_spp_map, am_spp_map, by = 'loiczid')%>%
    as.data.frame()

  return(spp_map)
}

### Plot Species Map function
### uses ggplot to create a formatted map of species ranges, including
### Aquamaps, IUCN, and overlapped ranges; returns the plot object.
plot_rangemap <- function(spp) {
  map <- get_spp_map(spp)

  am_map <- map %>%
    filter(!is.na(am_prob))

  r_am_spp  <-  subs(loiczid_raster, 
                     am_map[ , c('loiczid', 'am_prob')], 
                     by = 'loiczid', 
                     which = 'am_prob', 
                     subsWithNA = TRUE)
  r_am_spp[!is.na(r_am_spp)] <- 1

  iucn_map <- map %>%
    mutate(iucn_pres = iucn_area > 0)
  r_iucn_spp <- subs(loiczid_raster, 
                     iucn_map[ , c('loiczid', 'iucn_pres')], 
                     by = 'loiczid', 
                     which = 'iucn_pres', 
                     subsWithNA = TRUE)

  spp_pts <- as.data.frame(rasterToPoints(r_iucn_spp)) %>%
    full_join(as.data.frame(rasterToPoints(r_am_spp)),
              by = c('x', 'y')) %>%
    mutate(presence = ifelse(am_prob & is.na(iucn_pres), 'am',
                             ifelse(iucn_pres & is.na(am_prob), 'iucn',
                                    ifelse(iucn_pres & am_prob, 'both', NA))))

  spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
    ggtheme_map + 
    geom_raster(aes(fill = presence), alpha = .5) +
    scale_fill_manual(values = c('iucn' = 'blue', 'am' = 'red', 'both' = 'mediumorchid')) +
    borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
    scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
    scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
    labs(title = spp, x = NULL, y = NULL) 
}

```
***

```{r create_species_list_w_area, echo = FALSE, eval = FALSE, warning = FALSE, message = FALSE}

spp_list <- spp_list_base %>%
              mutate(total_area = NA,
                     am_area    = NA,
                     iucn_area  = NA,
                     perc       = NA,
                     am_perc    = NA,
                     iucn_perc  = NA,
                     sm_perc    = NA, #filled with % of smaller range that is within the larger range
                     sm_range   = NA) #what range is smaller

#there's likely a better way to do this than a for loop (lapply or sapply?)
for (i in 1:nrow(spp_list)){
  spp     <- spp_list[i, ]$sciname #species name
  spp_map <- get_spp_map(spp) #get species map

###get just the aquamaps cells
  am_spp_map <- spp_map %>%
    filter(!is.na(am_prob))%>%
    dplyr::select(loiczid, am_sid, am_prob) %>% 
      ### add in am_prob when we want to look at probability
    unique()

  if(any(duplicated(am_spp_map$loiczid))){
    am_spp_map <- am_spp_map %>%
      dplyr::select(loiczid, am_sid) %>% 
      unique()
    }

### rasterize aquamaps map
  r_am_spp <- subs(loiczid_raster, 
                   am_spp_map[ , c('loiczid', 'am_sid')], 
                   by = 'loiczid', 
                   which = 'am_sid', 
                   subsWithNA = TRUE)

  spp_list[i, ]$am_area <- area(r_am_spp, na.rm = TRUE)%>%
              cellStats(., stat='sum')

###  get the iucn cells
  iucn_map <- spp_map %>%
    mutate(iucn_pres = iucn_area > 0)%>%
      dplyr::select(loiczid, iucn_area, am_sid, am_prob, iucn_pres)%>%
        unique()

  #TODO: incorporate the iucn_area (area of cell that species overlaps with...I think thats what it is)

###this is incorporated because there are some duplications for loiczids with IUCN cells. I'm guessing this is because of the parent/subpopulations?

   if(TRUE %in% duplicated(iucn_map$loiczid))next()

### rasterize IUCN map
  r_iucn_spp <- subs(loiczid_raster, 
                     iucn_map[ , c('loiczid', 'iucn_pres')], 
                     by = 'loiczid', 
                     which = 'iucn_pres', 
                     subsWithNA = TRUE)

  spp_list[i, ]$iucn_area <- area(r_iucn_spp, na.rm = TRUE)%>%
                cellStats(., stat='sum')

### Get total area (IUCN and AM combined)
  allCells <- spp_map%>%
                dplyr::select(loiczid)%>%
                  mutate(value=1)%>%
                  unique()

  fullRange <- subs(loiczid_raster, 
                    allCells[ , c('loiczid', 'value')], 
                    by = 'loiczid', 
                    which = 'value', 
                    subsWithNA = TRUE)

  spp_list[i, ]$total_area <- area(fullRange, na.rm = TRUE)%>%
                              cellStats(., stat='sum')

### get percent overlap of map

   #iucn map cells
  cells_iucn<-iucn_map %>%
    filter(iucn_pres == TRUE) %>%
    .$loiczid
  cells_am <- am_spp_map %>%
    filter(!is.na(am_sid)) %>%
    .$loiczid

  cells_total <- unique(spp_map$loiczid)
  cells_overlap <- intersect(cells_iucn, cells_am)

  #I THINK THIS IS A DUPLICATE? r_iucn_spp raster created above... (JA 10/12/15)
#    r_iucn_spp <- subs(loiczid_raster, 
#                       iucn_map[ , c('loiczid', 'iucn_pres')], 
#                      by = 'loiczid', 
#                       which = 'iucn_pres', 
#                       subsWithNA = TRUE)
#   
  #percent overlap
  perc <- (length(cells_overlap) / length(cells_total)) * 100
  spp_list[i, ]$perc <- perc

  am_only <- setdiff(cells_am, cells_iucn)
  iucn_only <- setdiff(cells_iucn, cells_am)

  perc_am <- (length(am_only) / length(cells_total)) * 100
  perc_iucn <- (length(iucn_only) / length(cells_total)) * 100

  spp_list[i, ]$am_perc <- perc_am
  spp_list[i, ]$iucn_perc <- perc_iucn

  spp_list[i, ]$sm_range <- ifelse(length(cells_am)<length(cells_iucn), 'AM', 'IUCN') 
    ### what dataset has the smaller range
  spp_list[i, ]$sm_perc <- length(cells_overlap)/min(length(cells_am), length(cells_iucn)) * 100 
    ### this is the total percent of smaller range encompassed in the larger range

  cat(sprintf('%s: For species %s, %.2f%% of both datasets overlap, %.2f%% is just AquaMaps and %.2f%% is just IUCN\n', 
              i, spp, perc, perc_am, perc_iucn))
  }

# for Conus arenatus, the IUCN map gives 4 duplicate cells, but each of the entries has a different iucn_area...which doesn't make sense

#row 985 for Acanthurus leucopareius - AM is giving duplicate loiczids but with different am_prob...which is so strange. Looked
# at the raw data and it seems to be in the raw data... (JA on 8/31/15)

#remove non IUCN species and log total area (for use later on)
spp_list <- spp_list %>%
            filter(spp_group != 'hagfishes' & spp_group != 'non-homalopsids') %>%
              mutate(log_total_area = log(total_area))

write.csv(spp_list, file='spp_list_w_area_percent_noextinct.csv')
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# look at number of species per group

spp_list <- read.csv('spp_list_w_area_percent_noextinct.csv')

spp_df <- spp_list %>%
  mutate(diff = iucn_area - am_area)

@@

ggsave('figures/sp_groups_barchart.png', 
       plot = barchart_sp_groups, 
       width = 8, height = 4, dpi = 100)

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}

@@

ggsave('figures/CompareTotalArea_AM-IUCN_scatter.png', 
       plot = scatter_am_iucn_area, 
       width = 8, height = 4, dpi = 100)

@@

There are large differences so to get a better sense, here is the same comparison but with area log transformed.
``` {r echo = FALSE, warning = FALSE, message = FALSE}

@@

ggsave('figures/CompareTotalAreaLOG_AM-IUCN_scatter.png', 
       plot = scatter_am_iucn_area_log, 
       width = 8, height = 4, dpi = 100)

@@

Data split into species groups and by IUCN category
```{r, echo = FALSE, warning = FALSE, message = FALSE}

@@

ggsave('figures/CompareTotalAreaLOG_AM-IUCN_spgroups_scatter.png', 
       plot = scatter_am_iucn_alog_gp, 
       width = 8, height = 6, dpi = 100)

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}

@@

print(spp_map)
ggsave(plot = spp_map,
       filename = sprintf('figures/rangemap_%s.png', spp),
       width = 8, height = 4, dpi = 200)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

@@

```{r, echo = FALSE, warning = FALSE, message = FALSE}

@@

ggsave('figures/Overlap-PercTotalRange.png')

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE}

@@

ggsave('figures/Overlap-PercTotalRange_spGroups.png')
print(hist_overlap_pct_gp)

@@

###(4) Looking at the percent of smaller species range that is within the larger range.

For each range

```{r pct small within large, echo = FALSE, warning = FALSE}

spp_df4 <- spp_list %>%
  dplyr::select(sciname, sm_perc, sm_range) %>%
  unique()

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = 'Dataset with smaller range', labels = c('AquaMaps', 'IUCN')) + 
  labs(title = 'Percent of smaller range within larger range',
       x = 'Percent of smaller range that falls within the larger range', 
       y = sprintf('Proportion of all species (n = %s)', nrow(spp_df4)))

ggsave('figures/PercSmallRangeInLarger.png', width = 6, height = 4)

print(hist_sm_within_lg)

Exploring possible correlation between total species range area and percent overlap of smaller range within the larger:
``` {r pct overlap vs area, echo = FALSE, warning = FALSE, message = FALSE}

scatter_area_vs_overlap <- ggplot(spp_list, aes(sm_perc, total_area, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km^2)', 
       x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

lm_eqn <- function(df){
  m  <- lm(total_area ~ sm_perc, spp_list)
  eq <- substitute(italic('y') == a + b %.% italic('x')*","~~italic(r)^2~"="~r2, 
        list(a  = format(coef(m)[1], digits = 2), 
             b  = format(coef(m)[2], digits = 2), 
             r2 = format(summary(m)$r.squared, digits = 3)))
  as.character(as.expression(eq))             
}

scatter_area_vs_overlap <- scatter_area_vs_overlap + 
  geom_text(x = 40, y = 300000000, label = lm_eqn(spp_list), parse = TRUE)

print(scatter_area_vs_overlap)

# summary(lm(total_area ~ sm_perc, spp_list))

ggsave('figures/AreavsPercOverlap.png', width = 6, height = 4)

``` {r pct overlap vs ratio of area, echo = FALSE, warning = FALSE, message = FALSE}

spp_list <- spp_list %>%
  mutate(area_ratio = am_area / iucn_area,
         area_ratio = ifelse(area_ratio > 1, 1 / area_ratio, area_ratio),
         area_ratio = area_ratio * 100)

area_ratio_quartiles <- quantile(spp_list$area_ratio, na.rm = TRUE)[2:4]
pct_overlap_quartiles <- quantile(spp_list$sm_perc, na.rm = TRUE)[2:4]

spp_quads <- c('q1' = 'Kajikia albida', 
               'q2' = 'Galaxea cryptoramosa', 
               'q4' = 'Acanthurus nigroris', 
               'q4a' = 'Metanephrops australiensis', 
               'q3' = 'Conus magnificus')

spp_list_label <- spp_list %>%
  filter(sciname %in% spp_quads) %>%
  separate(sciname, c('g', 's'), sep = ' ', remove = FALSE) %>%
  mutate(sciname_label = paste(str_sub(g, 1, 1), '. ', s, sep = '')) %>%
  select(-g, -s)

lm_eqn2 <- function(df){
  m  <- lm(area_ratio ~ sm_perc, spp_list)
  eq <- substitute(italic('y') == a + b %.% italic('x')*","~~italic(r)^2~"="~r2, 
        list(a  = format(coef(m)[1], digits = 2), 
             b  = format(coef(m)[2], digits = 2), 
             r2 = format(summary(m)$r.squared, digits = 3)))
  as.character(as.expression(eq))             
}

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = log_total_area)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point() + 
  geom_vline(x = area_ratio_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'tomato2') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'tomato2',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap + 
  geom_text(x = 40, y = 5, label = lm_eqn2(spp_list), parse = TRUE, color = 'tomato2')

ggsave('figures/AreaRatiovsPercOverlap.png', width = 6, height = 4)

print(scatter_arearatio_vs_overlap)

# scatter_arearatio_vs_overlap2 <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = sm_range)) + 
#   ggtheme_plot + 
#   geom_point() + 
#   geom_vline(x = area_ratio_quartiles,
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles,
#              color = 'grey50', linetype = 'dashed') +
#   geom_point(data = spp_list_label, color = 'grey30') + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5) +
#   labs(x = 'Range area ratio (smaller:larger) * 100%', 
#        y = 'Percent overlap smaller range in larger', 
#        title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')
# 
# scatter_arearatio_vs_overlap2 <- scatter_arearatio_vs_overlap2 + 
#   geom_text(x = 40, y = 5, label = lm_eqn2(spp_list), parse = TRUE, color = 'grey30')
# 
# ggsave('figures/AreaRatiovsPercOverlap2.png', 
#        width = 6, height = 4)
# 
# print(scatter_arearatio_vs_overlap2)

# summary(lm(area_ratio ~ sm_perc, spp_list))

``` {r examine_spp_gp_quartiles, echo = FALSE}

spp_list1 <- spp_list %>%
  group_by(spp_group) %>%
  mutate(n_spp = n()) %>%
  ungroup()

spp_gp_q1 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio, na.rm = TRUE) & sm_perc >= median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q1 = n())
spp_gp_q2 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio, na.rm = TRUE) & sm_perc >= median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q2 = n())
spp_gp_q4 <- spp_list1 %>%
  filter(area_ratio < median(area_ratio, na.rm = TRUE) & sm_perc < median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q4 = n())
spp_gp_q3 <- spp_list1 %>%
  filter(area_ratio >= median(area_ratio, na.rm = TRUE) & sm_perc < median(sm_perc, na.rm = TRUE)) %>%
  group_by(spp_group) %>%
  summarize(n_spp = first(n_spp),
            n_spp_q3 = n())

spp_gp_quadrants <- spp_gp_q1 %>%
  left_join(spp_gp_q2, by = c('spp_group', 'n_spp')) %>%
  left_join(spp_gp_q3, by = c('spp_group', 'n_spp')) %>%
  left_join(spp_gp_q4, by = c('spp_group', 'n_spp')) %>%
  gather(quad, n_quad, n_spp_q1, n_spp_q2, n_spp_q3, n_spp_q4) %>% 
  mutate(quad = str_replace(quad, 'n_spp_', ''),
         pct_quad = n_quad/n_spp)

spp_gp_quadrants <- spp_gp_quadrants %>%
  left_join(spp_gp_quadrants %>% 
              filter(quad == 'q1') %>%
              select(spp_group, pct_q1 = pct_quad),
            by = 'spp_group')

barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group = reorder(spp_group, pct_q1)), 
                                aes(x = spp_group, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1' = 'green4', 'q2' = 'greenyellow', 'q3' = 'orange', 'q4' = 'orangered2')) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0) +
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Relative number of species per quadrant', 
       title = 'Species group breakdown by quadrant')

print(barchart_spp_gp_quads)
ggsave(plot = barchart_spp_gp_quads,
       filename = 'figures/barchart_spp_gp_quads.png',
       width = 6, height = 4)

```{r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE}

for (i in 1:length(spp_quads)) { # i = 1
  spp <- spp_quads[i]
  spp_quad <- names(spp_quads)[i]

  spp_plot <- plot_rangemap(spp)
  print(spp_plot)

  ggsave(filename = sprintf('figures/rangemap_%s_%s.png', spp_quad, str_replace(tolower(spp), ' ', '_')),
         plot = spp_plot, 
         width = 8, height = 4, dpi = 200)
}

Percent coverage as function of area, clipping to area below 100,000,000 km^2^:

``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# remove all area greater than 100000000

spp_small_df <- spp_list %>% 
  filter(total_area < 100000000)

scatter_a_vs_overlap_small <- ggplot(spp_small_df, aes(total_area, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  labs(x = 'Total Area (km^2)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

ggsave('figures/AreavsPercOverlap_2.png', width = 6, height = 4)

print(scatter_a_vs_overlap_small)

Percent coverage as a function of log(area)
``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# Log transform area

scatter_logA_vs_overlap <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  labs(x = 'Log(total Area)', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range (log) vs. percent overlap of \nsmaller range with larger')

ggsave('figures/AreavsPercOverlap_log.png', width=6, height=4)

print(scatter_logA_vs_overlap)
```

``` {r examine regressions, eval = FALSE, echo = FALSE, warning = FALSE}

A_mdl      <- lm(total_area ~ sm_perc, spp_list)
#                Estimate Std. Error t value Pr(>|t|)    
#   (Intercept) -16285410    2858388  -5.697  1.4e-08 ***
#   sm_perc        533370      41702  12.790  < 2e-16 ***
#   Multiple R-squared:  0.07732,	Adjusted R-squared:  0.07685

A_mdl_clip <- lm(total_area ~ sm_perc, l)
#               Estimate Std. Error t value Pr(>|t|)    
#   (Intercept)  2993939     810560   3.694 0.000227 ***
#   sm_perc       153523      11952  12.844  < 2e-16 ***
#   Multiple R-squared:  0.07955,	Adjusted R-squared:  0.07906 
logA_mdl   <- lm(log_total_area ~ sm_perc, spp_list)
#                Estimate Std. Error t value Pr(>|t|)    
#   (Intercept) 14.775325   0.096446   153.2   <2e-16 ***
#   sm_perc      0.016885   0.001407    12.0   <2e-16 ***
#   Multiple R-squared:  0.0687,	Adjusted R-squared:  0.06823 
```

###(5) Examining data broken into quartiles to see if smaller patterns emerge
``` {r examine quartiles, eval = FALSE, echo = FALSE, warning = FALSE}

summary(spp_list$total_area)
#      Min.   1st Qu.    Median      Mean   3rd Qu.      Max.
#     55760   3032000  10790000  18500000  22670000 372100000

qtiles <- c(quantile(spp_list$total_area, 0.0, na.rm = TRUE),
            quantile(spp_list$total_area, .25, na.rm = TRUE),
            quantile(spp_list$total_area, .50, na.rm = TRUE),
            quantile(spp_list$total_area, .75, na.rm = TRUE),
            quantile(spp_list$total_area, 1.0, na.rm = TRUE))

qtile_mdl <- list()
for (i in 1:4) {
  area_clip <- spp_list %>%
    filter(total_area >= qtiles[i] & total_area < qtiles[i + 1])
  qtile_mdl[[i]] <- lm(total_area ~ sm_perc, area_clip)

  qtile_plot <- ggplot(area_clip, aes(total_area, sm_perc, color = sm_range)) + 
    ggtheme_plot + 
    geom_point() + 
    labs(x = 'Total Area (km^2)', 
         y = 'Percent overlap smaller range in larger', 
         title = sprintf('Size of species range (quartile %s) vs. percent overlap of \nsmaller range with larger', i))

  ggsave(paste0('figures/AreavsOverlap_quantile_', i, '.png'), width = 6, height = 4)
  print(qtile_plot)
}
# First quartile regression:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept)  1423917     101053  14.091   <2e-16 ***
#     sm_perc         1188       1555   0.764    0.445    
#     Multiple R-squared:  0.001198,	Adjusted R-squared:  -0.0008529 
#
# second quartile:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept)  7011028     275943  25.408  < 2e-16 ***
#     sm_perc       -17725       4228  -4.192 3.28e-05 ***
#     Multiple R-squared:  0.03483,	Adjusted R-squared:  0.03285 
#
# third quartile:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept) 12813279     507856  25.230  < 2e-16 ***
#     sm_perc        64710       8159   7.931  1.5e-14 ***
#     Multiple R-squared:  0.1146,	Adjusted R-squared:  0.1128 
#
# fourth quartile:
#                   Estimate Std. Error t value Pr(>|t|)    
#     (Intercept) -101736351   17259544  -5.894 7.04e-09 ***
#     sm_perc        1911286     215216   8.881  < 2e-16 ***
#     Multiple R-squared:  0.1399,	Adjusted R-squared:  0.1381

# This avenue of exploration looks like a bust.

###(6) Looking at how many and what type (threat category) of species we are losing by setting a threshold of 0.4 on the aquamaps species

This looks at the full Aquamaps dataset... should it be limited to species with valid categories?

```{r aquamaps probabilities, warning = FALSE, message = FALSE, cache.lazy = FALSE, cache = TRUE, verbose = FALSE, results = FALSE, echo = FALSE, include = FALSE}

# get spp_cells raw data for aquamaps
dir_aquamaps <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO/raw/aquamaps_2014')

spp_cells <- read_csv(file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), 
                      col_types = '_ccd__', progress = TRUE)

# spp_cells <- fread(input = file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), verbose = FALSE, header = TRUE, 
#                   colClasses = c('NULL', 'character', NA, 'numeric', 'NULL', 'NULL')) #1 min 49 sec

setnames(spp_cells, c('SpeciesID', 'CsquareCode', 'probability'), c('species_id', 'csquare_code', 'probability'))

spp_cells <- as.data.frame(spp_cells)

spp <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
             verbose = FALSE, header = TRUE) %>%
  mutate(sciname = paste(Genus, Species, sep = ' '),
         iucn_code = str_replace(toupper(iucn_code), 'LR/', ''),
         iucn_code = str_replace(iucn_code, 'N.E.', 'NE'))
spp <- as.data.frame(spp)

```{r hist_lost_area_w_threshold, warning = FALSE, message = FALSE, echo = FALSE}

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  left_join(spp, by = c('species_id' = 'SPECIESID')) %>%
  mutate(prop = cells_0.4 / cells_t,
         iucn_code = ifelse(is.na(iucn_code), 'NE', iucn_code)) %>%
  filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')) %>%
  # filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

iucn_levels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')
spp_df3 <- spp_df3 %>% 
  transform(iucn_code = factor(iucn_code, levels = iucn_levels))

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  scale_fill_manual(values = c('LC' = 'green4', 
                               'NT' = 'greenyellow', 
                               'VU' = 'orange', 
                               'EN' = 'orangered2', 
                               'CR' = 'red4',
                               'DD' = 'grey60',
                               'NE' = 'grey70')) +
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', 
       y = sprintf('Proportion of all aquamaps species (n = %s)', nrow(spp_df3)))

ggsave('figures/AMThreshold.png', width = 6, height = 4)

print(hist_area_vs_threshold)

spp_df3a <- spp_df3 %>% 
  filter(!iucn_code %in% c('DD', 'NE'))

hist_area_vs_threshold2 <- ggplot(spp_df3a, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  scale_fill_manual(values = c('LC' = 'green4', 
                               'NT' = 'greenyellow', 
                               'VU' = 'orange', 
                               'EN' = 'orangered2', 
                               'CR' = 'red4',
                               'DD' = 'grey60',
                               'NE' = 'grey70')) +
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', 
       y = sprintf('Proportion of all aquamaps species (n = %s)', nrow(spp_df3a)))

ggsave('figures/AMThreshold2.png', width = 6, height = 4)

print(hist_area_vs_threshold2)

```{r scatter_lost_area_w_threshold, warning = FALSE, message = FALSE, echo = FALSE}

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat

>>>>

``` {r}
spp_df <- spp_list %>%
  mutate(diff = iucn_area - am_area)

@@

```{r number of spp per group, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

### look at number of species per group

@@

ggsave(file.path(dir_fig, 'sp_groups_barchart.png'), 
       plot = barchart_sp_groups, 
       width = 8, height = 4, dpi = 100)

@@

```{r area compare 1, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

@@

ggsave(file.path(dir_fig, 'CompareTotalArea_AM-IUCN_scatter.png'), 
       plot = scatter_am_iucn_area, 
       width = 8, height = 4, dpi = 100)

@@

There are large differences so to get a better sense, here is the same comparison but with area log transformed.
``` {r area compare log, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

ggsave(file.path(dir_fig, 'CompareTotalAreaLOG_AM-IUCN_scatter.png'), 
       plot = scatter_am_iucn_area_log, 
       width = 8, height = 4, dpi = 100)

@@

Data split into species groups and by IUCN category
```{r area compare by spp group, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

ggsave(file.path(dir_fig, 'CompareTotalAreaLOG_AM-IUCN_spgroups_scatter.png'), 
       plot = scatter_am_iucn_alog_gp, 
       width = 8, height = 6, dpi = 100)

@@

```{r map chromis alpha, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

@@

print(spp_map)
ggsave(plot = spp_map,
       filename = file.path(dir_fig, sprintf('rangemap_%s.png', spp)),
       width = 8, height = 4, dpi = 200)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}

@@

```{r hist overlap vs total range, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

ggsave(file.path(dir_fig, 'Overlap-PercTotalRange.png'))

@@

```{r facet by spp, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}

@@

ggsave(file.path(dir_fig, 'Overlap-PercTotalRange_spGroups.png'))
print(hist_overlap_pct_gp)

@@
